Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-27 background.tex
further shrinkage to bg figures, cleanup/reorg and compression of bg text

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.55,xscale=0.55]
%uncomment if require: \path (0,403); %set diagram left start at 0, and has height of 403

%Shape: Rectangle [id:dp9406932326447903] 
\draw   (33,133) -- (123,133) -- (123,284) -- (33,284) -- cycle ;
%Shape: Rectangle [id:dp7682116051394419] 
\draw   (166,133) -- (256,133) -- (256,284) -- (166,284) -- cycle ;
%Shape: Rectangle [id:dp07376733516930423] 
\draw   (295,133) -- (385,133) -- (385,284) -- (295,284) -- cycle ;
%Shape: Rectangle [id:dp7106629147891306] 
\draw   (427,133) -- (517,133) -- (517,284) -- (427,284) -- cycle ;
%Shape: Rectangle [id:dp1342906826203598] 
\draw   (115,46) -- (185,46) -- (185,86) -- (115,86) -- cycle ;
%Shape: Rectangle [id:dp08072439007207399] 
\draw   (185,46) -- (255,46) -- (255,86) -- (185,86) -- cycle ;
%Shape: Rectangle [id:dp8282933603833216] 
\draw   (255,46) -- (325,46) -- (325,86) -- (255,86) -- cycle ;
%Shape: Rectangle [id:dp8416255090865326] 
\draw   (325,46) -- (395,46) -- (395,86) -- (325,86) -- cycle ;
%Shape: Rectangle [id:dp22911958218529804] 
\draw   (395,46) -- (465,46) -- (465,86) -- (395,86) -- cycle ;
%Shape: Rectangle [id:dp33732955213990756] 
\draw   (464.5,46) -- (555.5,46) -- (555.5,86) -- (464.5,86) -- cycle ;
%Curve Lines [id:da6688730321284677] 
\draw    (211.5,88) .. controls (-36.44,90.94) and (15.75,182.24) .. (30.17,205.65) ;
\draw [shift={(31,207)}, rotate = 238.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp6044747978986409] 
\draw   (33,183) -- (123,183) -- (123,223) -- (33,223) -- cycle ;
%Curve Lines [id:da8867562998134584] 
\draw    (289.5,87) .. controls (115.15,102.76) and (137.77,167.03) .. (163.33,189.98) ;
\draw [shift={(164.5,191)}, rotate = 220.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp8861153167245819] 
\draw   (166,169) -- (256,169) -- (256,209) -- (166,209) -- cycle ;
%Shape: Rectangle [id:dp05979844200848894] 
\draw  [dash pattern={on 4.5pt off 4.5pt}] (295,148) -- (385,148) -- (385,188) -- (295,188) -- cycle ;
%Shape: Rectangle [id:dp17403733244247843] 
\draw   (427,204) -- (517,204) -- (517,244) -- (427,244) -- cycle ;
%Curve Lines [id:da20581124900573955] 
\draw [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ] [dash pattern={on 4.5pt off 4.5pt}]  (357,89) .. controls (262.92,102.72) and (272.56,140.45) .. (291.81,165.48) ;
\draw [shift={(293,167)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da07589611603635316] 
\draw    (425.5,87) .. controls (380.42,111.5) and (406.41,194.58) .. (422.53,216.72) ;
\draw [shift={(423.5,218)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da6852167546673456] 
\draw    (497.5,87) .. controls (564.48,83.06) and (567.43,193.61) .. (570.37,231.33) ;
\draw [shift={(570.5,233)}, rotate = 265.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da017521077423665377] 
\draw    (36.5,330) .. controls (-15.71,340.84) and (16.98,314.8) .. (32.31,285.35) ;
\draw [shift={(33,284)}, rotate = 116.57] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da6750042432035159] 
\draw    (122,207) .. controls (128.9,246.4) and (131.91,261.54) .. (159.71,281.1) ;
\draw [shift={(161,282)}, rotate = 214.59] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da14788158682113428] 
\draw    (257,189) .. controls (258.95,228) and (277.06,262.25) .. (293.72,282.47) ;
\draw [shift={(295,284)}, rotate = 229.64] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da3297977595767969] 
\draw    (386,169) .. controls (392.86,218) and (408.36,262.2) .. (425.92,282.77) ;
\draw [shift={(427,284)}, rotate = 228.01] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da7710391961792686] 
\draw    (517,221) .. controls (535.13,236.68) and (527.81,247.56) .. (555.75,240.46) ;
\draw [shift={(557.5,240)}, rotate = 165.07] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp1535611125971872] 
\draw   (511,313) -- (581,313) -- (581,353) -- (511,353) -- cycle ;
%Shape: Rectangle [id:dp05427947476226991] 
\draw   (95,313) -- (136.5,313) -- (136.5,353) -- (95,353) -- cycle ;
%Shape: Rectangle [id:dp10511138743752002] 
\draw   (136.5,313) -- (265.5,313) -- (265.5,353) -- (136.5,353) -- cycle ;
%Shape: Rectangle [id:dp8917680919534263] 
\draw   (265.5,313) -- (305,313) -- (305,353) -- (265.5,353) -- cycle ;
\draw   (560,243.5) .. controls (560,238.25) and (564.25,234) .. (569.5,234) .. controls (574.75,234) and (579,238.25) .. (579,243.5) .. controls (579,248.75) and (574.75,253) .. (569.5,253) .. controls (564.25,253) and (560,248.75) .. (560,243.5) -- cycle ; \draw   (560,243.5) -- (579,243.5) ; \draw   (569.5,234) -- (569.5,253) ;
%Curve Lines [id:da9312068388962929] 
\draw    (569,258) .. controls (585.66,269.76) and (587.91,278.64) .. (566.35,295.93) ;
\draw [shift={(565,297)}, rotate = 321.95] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

% Text Node
\draw (21,20) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (569,375) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Physical}};
% Text Node
\draw (102,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 63}};
% Text Node
\draw (173,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 48}};
% Text Node
\draw (243,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 39}};
% Text Node
\draw (312,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 30}};
% Text Node
\draw (383,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 21}};
% Text Node
\draw (452,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% Text Node
\draw (547,16) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 0}};
% Text Node
\draw (126,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Sign}};
% Text Node
\draw (188,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (257,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (328,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (398,55) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (471,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (87,52) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize va}};
% Text Node
\draw (46,192) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (178,178) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (309,156) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (442,212) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (59,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize root}};
% Text Node
\draw (471,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize page}};
% Text Node
\draw (517,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (131,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 51}};
% Text Node
\draw (256,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% Text Node
\draw (161,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}

          \caption{x86-64 page table lookups.}
        \label{fig:pagetables}
    \end{subfigure}
    \vfill
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\scriptsize]
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /* Check if entry is populated, allocate if needed(* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)(*\label{line:after_concluding_qfrac1C}*)*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    pte_initialize(entry); // Alloc empty table for next level (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)(* \label{line:call_to_pte_initializeC} *)
    entry->present = 1; // Mark valid (*\label{line:install_new_entryC}*)
    /* Key invariant for child access now holds */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\end{minipage}
\hfill\hfill
\begin{minipage}{.4\textwidth}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\scriptsize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{A kernel implementation of x86-64 page table lookups.}
    \label{fig:pagetablescode}
    \vspace{-1em}
\end{figure}

\todo[inline]{Colin says: the changes in the next couple paragraphs seem to have deleted critical information about how the hardware
translates addresses.}
On the x86-64 architecture, the \textsc{MMU}'s address translation uses a sparse hierarchical set of tables:
\emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} (based on Figure 5-17 of the 
AMD64 architecture manual~\cite{amd64_manual_vol2}\footnote{While x86 up through its 32-bit incarnation were due to Intel,
the x86-64 architecture as a 64-bit extension to x86 was originally due to AMD. As a result, it is sometimes also referred to as the \texttt{amd64} architecture.})
\replace{shows,
address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table.
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved.
On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
}{
shows a \emph{page-table walk} for a virtual address. The page tables are managed by the OS, typically by a \emph{virtual memory manager} (VMM).\footnote{Not to be confused with Virtual 
Machine Monitor. We focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
Typically each process has its own 
page table, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (the start of the L4 table shown as $\mathsf{root}$ in Figure \ref{fig:pagetables}) in a specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.
}

>>>>

% \tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

% \begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.55,xscale=0.55]
% %uncomment if require: \path (0,403); %set diagram left start at 0, and has height of 403

% %Shape: Rectangle [id:dp9406932326447903] 
% \draw   (33,133) -- (123,133) -- (123,284) -- (33,284) -- cycle ;
% %Shape: Rectangle [id:dp7682116051394419] 
% \draw   (166,133) -- (256,133) -- (256,284) -- (166,284) -- cycle ;
% %Shape: Rectangle [id:dp07376733516930423] 
% \draw   (295,133) -- (385,133) -- (385,284) -- (295,284) -- cycle ;
% %Shape: Rectangle [id:dp7106629147891306] 
% \draw   (427,133) -- (517,133) -- (517,284) -- (427,284) -- cycle ;
% %Shape: Rectangle [id:dp1342906826203598] 
% \draw   (115,46) -- (185,46) -- (185,86) -- (115,86) -- cycle ;
% %Shape: Rectangle [id:dp08072439007207399] 
% \draw   (185,46) -- (255,46) -- (255,86) -- (185,86) -- cycle ;
% %Shape: Rectangle [id:dp8282933603833216] 
% \draw   (255,46) -- (325,46) -- (325,86) -- (255,86) -- cycle ;
% %Shape: Rectangle [id:dp8416255090865326] 
% \draw   (325,46) -- (395,46) -- (395,86) -- (325,86) -- cycle ;
% %Shape: Rectangle [id:dp22911958218529804] 
% \draw   (395,46) -- (465,46) -- (465,86) -- (395,86) -- cycle ;
% %Shape: Rectangle [id:dp33732955213990756] 
% \draw   (464.5,46) -- (555.5,46) -- (555.5,86) -- (464.5,86) -- cycle ;
% %Curve Lines [id:da6688730321284677] 
% \draw    (211.5,88) .. controls (-36.44,90.94) and (15.75,182.24) .. (30.17,205.65) ;
% \draw [shift={(31,207)}, rotate = 238.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Shape: Rectangle [id:dp6044747978986409] 
% \draw   (33,183) -- (123,183) -- (123,223) -- (33,223) -- cycle ;
% %Curve Lines [id:da8867562998134584] 
% \draw    (289.5,87) .. controls (115.15,102.76) and (137.77,167.03) .. (163.33,189.98) ;
% \draw [shift={(164.5,191)}, rotate = 220.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Shape: Rectangle [id:dp8861153167245819] 
% \draw   (166,169) -- (256,169) -- (256,209) -- (166,209) -- cycle ;
% %Shape: Rectangle [id:dp05979844200848894] 
% \draw  [dash pattern={on 4.5pt off 4.5pt}] (295,148) -- (385,148) -- (385,188) -- (295,188) -- cycle ;
% %Shape: Rectangle [id:dp17403733244247843] 
% \draw   (427,204) -- (517,204) -- (517,244) -- (427,244) -- cycle ;
% %Curve Lines [id:da20581124900573955] 
% \draw [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ] [dash pattern={on 4.5pt off 4.5pt}]  (357,89) .. controls (262.92,102.72) and (272.56,140.45) .. (291.81,165.48) ;
% \draw [shift={(293,167)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da07589611603635316] 
% \draw    (425.5,87) .. controls (380.42,111.5) and (406.41,194.58) .. (422.53,216.72) ;
% \draw [shift={(423.5,218)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da6852167546673456] 
% \draw    (497.5,87) .. controls (564.48,83.06) and (567.43,193.61) .. (570.37,231.33) ;
% \draw [shift={(570.5,233)}, rotate = 265.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da017521077423665377] 
% \draw    (36.5,330) .. controls (-15.71,340.84) and (16.98,314.8) .. (32.31,285.35) ;
% \draw [shift={(33,284)}, rotate = 116.57] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da6750042432035159] 
% \draw    (122,207) .. controls (128.9,246.4) and (131.91,261.54) .. (159.71,281.1) ;
% \draw [shift={(161,282)}, rotate = 214.59] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da14788158682113428] 
% \draw    (257,189) .. controls (258.95,228) and (277.06,262.25) .. (293.72,282.47) ;
% \draw [shift={(295,284)}, rotate = 229.64] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da3297977595767969] 
% \draw    (386,169) .. controls (392.86,218) and (408.36,262.2) .. (425.92,282.77) ;
% \draw [shift={(427,284)}, rotate = 228.01] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Curve Lines [id:da7710391961792686] 
% \draw    (517,221) .. controls (535.13,236.68) and (527.81,247.56) .. (555.75,240.46) ;
% \draw [shift={(557.5,240)}, rotate = 165.07] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
% %Shape: Rectangle [id:dp1535611125971872] 
% \draw   (511,313) -- (581,313) -- (581,353) -- (511,353) -- cycle ;
% %Shape: Rectangle [id:dp05427947476226991] 
% \draw   (95,313) -- (136.5,313) -- (136.5,353) -- (95,353) -- cycle ;
% %Shape: Rectangle [id:dp10511138743752002] 
% \draw   (136.5,313) -- (265.5,313) -- (265.5,353) -- (136.5,353) -- cycle ;
% %Shape: Rectangle [id:dp8917680919534263] 
% \draw   (265.5,313) -- (305,313) -- (305,353) -- (265.5,353) -- cycle ;
% \draw   (560,243.5) .. controls (560,238.25) and (564.25,234) .. (569.5,234) .. controls (574.75,234) and (579,238.25) .. (579,243.5) .. controls (579,248.75) and (574.75,253) .. (569.5,253) .. controls (564.25,253) and (560,248.75) .. (560,243.5) -- cycle ; \draw   (560,243.5) -- (579,243.5) ; \draw   (569.5,234) -- (569.5,253) ;
% %Curve Lines [id:da9312068388962929] 
% \draw    (569,258) .. controls (585.66,269.76) and (587.91,278.64) .. (566.35,295.93) ;
% \draw [shift={(565,297)}, rotate = 321.95] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

% % Text Node
% \draw (21,20) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Virtual}};
% % Text Node
% \draw (569,375) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Physical}};
% % Text Node
% \draw (102,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 63}};
% % Text Node
% \draw (173,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 48}};
% % Text Node
% \draw (243,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 39}};
% % Text Node
% \draw (312,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 30}};
% % Text Node
% \draw (383,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 21}};
% % Text Node
% \draw (452,17) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% % Text Node
% \draw (547,16) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 0}};
% % Text Node
% \draw (126,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Sign}};
% % Text Node
% \draw (188,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Offset}};
% % Text Node
% \draw (257,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Offset}};
% % Text Node
% \draw (328,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Offset}};
% % Text Node
% \draw (398,55) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Offset}};
% % Text Node
% \draw (471,54) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Page Offset}};
% % Text Node
% \draw (87,52) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize va}};
% % Text Node
% \draw (46,192) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L4 Entry}};
% % Text Node
% \draw (178,178) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L3 Entry}};
% % Text Node
% \draw (309,156) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L2 Entry}};
% % Text Node
% \draw (442,212) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize L1 Entry}};
% % Text Node
% \draw (59,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize root}};
% % Text Node
% \draw (471,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize page}};
% % Text Node
% \draw (517,321) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 4K Data}};
% % Text Node
% \draw (131,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 51}};
% % Text Node
% \draw (256,358) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize 12}};
% % Text Node
% \draw (161,322) node [anchor=north west][inner sep=0.75pt]  [font=\scriptsize] [align=left] {{\scriptsize Base Addr}};

% \end{tikzpicture}

\includegraphics[width=\textwidth]{addr_translation.png}
\vspace{-1em}
          \caption{x86-64 \replace{page table lookups}{hardware address translation}.}
        \label{fig:pagetables}
    \end{subfigure}
    \vfill
\hspace{5em}
\begin{minipage}{.5\textwidth}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)},basicstyle=\scriptsize]
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  if (!entry->present) { // Check if entry needs initialization
    pte_initialize(entry); // Alloc empty table for next level(*\label{line:ptc_alloc}*)
    entry->present = 1; // Mark valid(*\label{line:ptc_mark}*)
  }
  /* Convert phys. addr. from PTE to virt. addr. for access */
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);(*\label{line:ptc_p2v}*)
  next = (pte_t *) next_virt_addr;
  return next;
}
\end{lstlisting}
\end{minipage}
\hfill\hfill
\begin{minipage}{.4\textwidth}
\hspace{3em}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}, basicstyle=\scriptsize]
pte_t *walkpgdir(pte_t *root, void *va){ 
  pte_t *l4_entry = &root[L4Offset(va)];
  pte_t *l3 = pte_next_table(l4_entry);
  pte_t *l3_entry = &l3[L3Offset(va)];
  pte_t *l2 = pte_next_table(l3_entry);
  pte_t *l2_entry = &l2[L2Offset(va)];  
  pte_t *l1 = pte_next_table(l2_entry);
  pte_t *l1_entry = &l1[L1Offset(va)];  
  return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{minipage}
%\end{framed}
    \vspace{-2.5em}
    \caption{Most of a software page table walk in C, as used to map new pages.}
    \label{fig:pagetablescode}
    \vspace{-1em}
\end{figure}

On the x86-64 architecture, the \textsc{MMU}'s address translation uses a sparse hierarchical set of tables:
\emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} (based on Figure 5-17 of the 
AMD64 architecture manual~\cite{amd64_manual_vol2}\footnote{While x86 up through its 32-bit incarnation were due to Intel,
the x86-64 architecture as a 64-bit extension to x86 was originally due to AMD. As a result, it is sometimes also referred to as the \texttt{amd64} architecture.})
shows,
address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into \replace{a}{successive} table\add{s} \add{of 512 8-byte entries (making each table 4KB in size)}.
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory \add{accessible to the running
program}, to which the low-order
\add{12} bits of the accessed virtual address are added to determine the actual physical address retrieved.
On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have explicit historical names,
but for brevity and consistency, we simply number them, in keeping with the newer 5th level. Our formalization only deals with 4-level page tables, but is straightforwardly extensible to 5.} 
The translation process or algorithm is \replace{sometimes}{often} referred to as a \emph{page-table walk}.
While Figure \ref{fig:pagetables} and most of our constants (how many levels, which virtual address bits index which
table levels) are specific to \replace{the}{common} x86-64 \replace{architecture}{CPUs},
\replace{
ARMv8 (a.k.a.\ texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.
RISC-V's \texttt{sv48} paging configuration~\cite[\S 4.5]{riscv-privileged} and AArch64's 4-level paging configuration
both split 64-bit virtual addresses at the same points to index into the respective tables, so most of what follows
is equally applicable to those architectures (the only difference is that page table entries place control bits of write access, etc.,
in different orders).
}{
  it is straightforward to adapt our approach to ARMv8/\texttt{aarch64}, RISC-V, or PowerPC, which use nearly-identical
  page table structures (only the order of flag bits in the entries differ),
  or to 5-level paging present in newer x86-64 or RISC-V designs (one step would be iterated an additional time).
}

The entries \add{(e.g., L4 Entry in Figure \ref{fig:pagetables})} of each table are 64 bits wide,
but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set),
and a range of additional bits which can be used to control caching, write-through, and more.
This paper will only consider the present bit (0).
